`r glue::glue('inferCNV results for {params$library_id} with {params$reference_name} reference')`

### 2025-05-07 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd
update density plots to have two ridgeplots instead at different cell groupings, and begin updating cell type scheme

# Define cell types based on ewing module
celltype_df <- readr::read_tsv(ewings_tsv) |>
  # Label cells as:
  # - tumor-immune: Cells labeled as tumor by cell-type-ewings, but immune by cell-type-consensus
  # - tumor-endo: Cells labeled as tumor by cell-type-ewings, but endothelial by cell-type-consensus
  # - tumor: Cells labeled as tumor by cell-type-ewings
  # - their existing ewing_annotation if none of the above
  dplyr::mutate(
    celltype = dplyr::case_when(
      ## tumor-immune
      consensus_annotation %in% immune_labels &
        stringr::str_starts(ewing_annotation, "tumor") ~ "tumor-immune",
      ## tumor-endo
      consensus_annotation %in% endo_labels &
        stringr::str_starts(ewing_annotation, "tumor") ~ "tumor-endo",
      ## tumor
      stringr::str_starts(ewing_annotation, "tumor") ~ "tumor",
      ## default
      .default = ewing_annotation
    )
  ) |>
  dplyr::select(barcodes, celltype)

umap_df <- umap_df |>
  dplyr::left_join(celltype_df)
```

@@

The figures below compare the number of CNVs per cell between reference and query cells.
Ideally, we'll see a bimodal distribution where reference cells are at/near zero, and there is a peak of higher values which may represent tumor cells.
In the density plots below, distributions are colored based on whether the cells are in the reference or not (aka, "unknown" cells).

```{r}
# First, construct data frame of total `has_cnv` counts, with other variables needed for plots
has_cnv_df <- infercnv_df |>
  dplyr::select(
    barcodes,
    starts_with("has_cnv_")
  ) |>
  tidyr::pivot_longer(
    starts_with("has_cnv_"),
    names_to = "chr",
    values_to = "cnv"
  ) |>
  # sum all values for each
  dplyr::group_by(barcodes) |>
  dplyr::summarize(total_cnv_per_cell = sum(cnv)) |>
  dplyr::ungroup() |>
  # bring back metadata
  dplyr::inner_join(
    infercnv_metadata_df,
    by = "barcodes"
  ) |>
  # join in SCE information
  dplyr::left_join(umap_df)
```

```{r fig.width = 10, fig.height = 5}
ggplot(has_cnv_df) +
  aes(x = total_cnv_per_cell, fill = cell_group) +
  geom_density(alpha = 0.5) +
  theme(legend.position = "bottom")
```

@@

* `tumor-endo`: Cells labeled as tumor by cell-type-ewings, but endothelial by `cell-type-consensus`
* `tumor-immune`: Cells labeled as tumor by cell-type-ewings, but immune by `cell-type-consensus`
* `tumor`: Cells labeled as tumor by `cell-type-ewings`
* If none of the above, their existing `cell-type-ewings` annotation

>>>>

# Read ewings cell types
celltype_df <- readr::read_tsv(ewings_tsv)
```

@@

The figures below compare the number of CNVs per cell between reference and query cells.
Ideally, we'll see a bimodal distribution where reference cells are at/near zero, and there is a peak of higher values which may represent tumor cells.

```{r}
# First, construct data frame of total `has_cnv` counts, with other variables needed for plots
has_cnv_df <- infercnv_df |>
  dplyr::select(
    barcodes,
    starts_with("has_cnv_")
  ) |>
  tidyr::pivot_longer(
    starts_with("has_cnv_"),
    names_to = "chr",
    values_to = "cnv"
  ) |>
  # sum all values for each
  dplyr::group_by(barcodes) |>
  dplyr::summarize(total_cnv_per_cell = sum(cnv)) |>
  dplyr::ungroup() |>
  # bring back metadata
  dplyr::inner_join(
    infercnv_metadata_df,
    by = "barcodes"
  ) |>
  # join in SCE information
  dplyr::left_join(celltype_df) |>
  dplyr::left_join(umap_df) |>
  # add cell type grouping column, with counts
  dplyr::mutate(
    celltype = dplyr::case_when(
      cell_group == "reference" ~ "reference cell",
      stringr::str_detect(ewing_annotation, "tumor") ~ "tumor cell",
      .default = "other query cell"
    )
  ) |>
  # add columns with cell type & group counts
  dplyr::add_count(celltype, name = "celltype_count") |>
  dplyr::add_count(cell_group, name = "cell_group_count") |>
  dplyr::mutate(
    celltype_n = glue::glue("{celltype} (n={celltype_count})"),
    cell_group_n = glue::glue("{cell_group} (n={cell_group_count})")
  )

# Set the celltype_n factor order based on label, not frequency
# We don't know the exact labels due to `(n=)`, so use grep
has_cnv_df$celltype_n <- as.factor(has_cnv_df$celltype_n)
present_values <- levels(has_cnv_df$celltype_n)
has_cnv_df$celltype_n <- forcats::fct_relevel(
  has_cnv_df$celltype_n,
  c(
    grep("reference", present_values, value = TRUE),
    grep("tumor", present_values, value = TRUE),
    grep("other", present_values, value = TRUE)
  )
)
```

Below, we show two ridge plots for the total detected CNVs across cell groups:

* On the left, distributions are colored based on whether the cells are in the reference or not (aka, "query" cells)
* On the right, distributions are colored based on a broad cell type category as follows:
  * `reference cell`: Cells in the normal reference
  * `tumor cell`: Cells labeled as tumor by the `cell-type-ewings` module
  * `other query cell`: All other cells

```{r fig.width = 12, fig.height = 5}
left_plot <- ggplot(has_cnv_df) +
  aes(y = forcats::fct_rev(cell_group_n), x = total_cnv_per_cell, fill = cell_group_n) +
  ggridges::geom_density_ridges2(alpha = 0.7, scale = 0.85) +
  labs(y = "Cell group (number of cells)") +
  theme(legend.position = "bottom")

right_plot <- ggplot(has_cnv_df) +
  aes(y = forcats::fct_rev(celltype_n), x = total_cnv_per_cell, fill = celltype_n) +
  ggridges::geom_density_ridges2(alpha = 0.7, scale = 0.85) +
  labs(y = "Cell type (number of cells)") +
  theme(legend.position = "bottom") +
  scale_fill_brewer(palette = "Dark2")

left_plot + right_plot
```

@@
